% Shortened appendix for Section~\ref{sec:glasgow} — probe settings + one table.
% The results table lives in probing_results_table.tex (same directory) and is
% \input below. Requires \usepackage{booktabs,multirow,bm} — standard amsmath.
% (The pre-k-fold long version is appendix_probing.tex; include only one — both
% define \label{app:probes}.)

\section{Probe Settings and Per-Dimension Results}
\label{app:probes}

\paragraph{Data and embeddings.}
The official Glasgow norms ($5{,}553$ entries, including the $871$
sense-disambiguated items such as \emph{apple (fruit)} / \emph{Apple (brand)},
embedded as the full string) and the Brysbaert concreteness ratings
($39{,}954$ words). Each word is embedded once per encoder with its standard
output pipeline: MiniLM (\texttt{all-MiniLM-L6-v2}, $384$-d) and EmbeddingGemma
(\texttt{embeddinggemma-300M}, $768$-d, \textsc{sts} prompt).

\paragraph{Evaluation protocol.}
Five-fold cross-validation (shuffled, seed $42$): each fold trains on $80\%$ and
validates on a disjoint $20\%$, so every word is scored exactly once; all probes
share the same folds. Cells report mean $\pm$ standard deviation over folds.
Activations are per-feature $z$-scored using each fold's training statistics.

\paragraph{Probe settings.}
\textsc{Ridge}: $\ell_2$ linear regression, $\alpha=1$. \textsc{SVR}:
support-vector regression, RBF kernel, $C=1$. \textsc{MLP}: one hidden layer
($256$, ReLU, dropout $0.1$), Adam (lr $10^{-3}$, weight decay $10^{-4}$), MSE
loss, batch size $64$, a fixed $100$ epochs with no early stopping (avoiding
best-checkpoint selection on the validation fold). \textsc{Mass-mean}: the
closed-form unit-normalised difference-of-means direction
$\bm{\mu}_{+}-\bm{\mu}_{-}$ of \citet{marks2023geometry} on a median split of
the target, with no covariance correction; because the projection
$s(\bm{x})=\hat{\bm{\theta}}^{\top}\tilde{\bm{x}}$ is uncalibrated, we fit a
univariate readout $y \approx a\,s(\bm{x})+c$ on the fold's training split and
report its held-out $R^2$, making the column directly comparable to the other
regression probes. \textsc{Logistic}: $\ell_2$ logistic regression ($C=1$, up to
$1000$ iterations) on a median split of the norm, reported as classification
accuracy. Ridge, SVR and the MLP report $R^2$.

\paragraph{Reading the table.}
$\Delta=$ SVR$-$ridge mean $R^2$ measures how much a nonlinear probe adds over
the best linear one: it is at most $0.10$ (median $0.05$) across all norms and
both encoders, and a single calibrated mass-mean direction alone recovers
$R^2$ $0.35$--$0.74$ --- the dimensions are largely linearly encoded.

\input{probing_results_table}
