% Appendix: probe protocol + per-dimension results for the psycholinguistic-
% norms probing study (Section~\ref{sec:glasgow}).
% Requires \usepackage{booktabs}. Uses \(\ell_2\) etc. — standard amsmath.

\section{Probe Protocol and Per-Dimension Results}
\label{app:probes}

Each norm is treated as an independent 1-D target recovered from a frozen
embedding by a probe; we compare probe families and embedding spaces on one
shared evaluation surface. Activations are extracted once per model and cached,
so re-runs only re-train probes.

\paragraph{Embedding spaces.}
Two sentence encoders are probed over the $4{,}682$ Glasgow words: MiniLM
(\texttt{all-MiniLM-L6-v2}, $384$-d, CLS pooling, hidden layers $\{0,3,6\}$)
and EmbeddingGemma (\texttt{google/embeddinggemma-300M}, $768$-d, mean pooling,
hidden layers $\{0,12,24\}$). Each extraction emits one pooled hidden vector per
word per layer; a probe is fit independently at every layer and we report the
best-layer validation score. The concreteness-only study (Brysbaert et al.,
$39{,}954$ words) probes the final pooled MiniLM embedding.

\paragraph{Targets.}
Norms are used as continuous regression targets. For the logistic probe only,
a target is binarised at its median (above/below), following the two-class
framing of \citet{marks2023geometry}. Items with a null target are excluded
from training but still scored. We require at least $N_{\min}=50$ valid targets.

\paragraph{Train/validation split.}
For each (space, target) we draw a single random $80/20$ train/validation split
(seed $=42$) and reuse the \emph{same} split across all probe families, so their
validation metrics are directly comparable. Inputs are cast to \texttt{float32}.
Unless noted, activations are per-feature $z$-scored using train-split
statistics.

\paragraph{Ridge probe.}
$\ell_2$-regularised linear regression on standardised activations,
$\alpha=1.0$. The learned direction $\hat{\bm{\theta}}$ is the weight vector;
each item is scored $s(\bm{x}) = \hat{\bm{\theta}}^{\top}\tilde{\bm{x}} + b$.

\paragraph{Lasso probe.}
$\ell_1$-regularised linear regression, $\alpha=0.01$ (up to $5000$
coordinate-descent iterations), used mainly for the sparse direction vectors it
exposes for downstream feature analysis.

\paragraph{SVR probe.}
Support-vector regression with an RBF kernel, $C=1.0$ — the nonlinear regressor;
it is the strongest probe for both small encoders. Scores are the SVR
predictions.

\paragraph{Mass-mean probe.}
The closed-form difference-of-class-means direction of \citet{marks2023geometry}
on standardised activations: with a median split of the target,
$\hat{\bm{\theta}}_{\mathrm{mm}} = \bm{\mu}_{+}-\bm{\mu}_{-}$ (unit-normalised),
scored by the projection $s(\bm{x}) = \hat{\bm{\theta}}_{\mathrm{mm}}^{\top}
\tilde{\bm{x}}$. Because the projection is uncalibrated, its natural metrics are
Pearson/Spearman correlation; where an $R^2$ is reported we fit a univariate
least-squares readout $y \approx a\,s(\bm{x}) + c$ on the training split and
evaluate it on validation. Standardising the activations makes this a
diagonal-covariance variant of the raw mass-mean probe; the full $\Sigma^{-1}$
correction, whose regularised form is subsumed by the ridge probe, is not
applied.

\paragraph{Logistic-regression probe.}
The discriminative counterpart to the mass-mean $\sigma(\cdot)$ probe. Each norm
is split at its median into two equal-frequency classes (concrete vs.\ abstract,
etc.); we fit $\ell_2$-regularised logistic regression (scikit-learn
\texttt{LogisticRegression}, inverse-regularisation $C=1.0$, L-BFGS, up to
$1000$ iterations, uniform class weights) on the same standardised activations
and $80/20$ split. The class direction is the fitted weight vector
$\hat{\bm{\theta}}$; items are scored by the decision value
$\hat{\bm{\theta}}^{\top}\tilde{\bm{x}} + b$. Because the target is binary, this
probe reports \emph{classification} metrics (accuracy, weighted $F_1$, ROC-AUC)
rather than $R^2$.

\paragraph{MLP probe.}
A nonlinear baseline: one hidden layer of width $256$ with ReLU and dropout
$0.1$, i.e. $\mathrm{Linear}(d,256)\!\to\!\mathrm{ReLU}\!\to\!\mathrm{Dropout}
(0.1)\!\to\!\mathrm{Linear}(256,1)$, trained with Adam (learning rate $10^{-3}$,
weight decay $10^{-4}$), mean-squared-error loss, batch size $64$, for up to
$100$ epochs with early stopping (patience $10$) on validation loss; the
best-validation checkpoint is restored. It has no explicit direction vector;
scores are the network's predictions. Training is seeded (seed $=42$).

\paragraph{Reported metrics.}
For regression probes we report the best-layer validation $R^2$ and Spearman
$\rho$; the store additionally records Pearson $r$, MSE, MAE and the train-split
values. The mass-mean probe is reported as $\rho$ (a correlation, not $R^2$).
The logistic probe is reported as validation accuracy and ROC-AUC. Non-finite
values from degenerate targets are recorded as null rather than reported.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{@{}ll@{}}
    \toprule
    \textbf{Setting} & \textbf{Value} \\
    \midrule
    \multicolumn{2}{@{}l}{\emph{Shared}} \\
    Train/validation split      & $80/20$ \\
    Random seed                 & $42$ \\
    Min.\ valid targets $N_{\min}$ & $50$ \\
    Standardisation             & per-feature $z$-score \\
    Input dtype                 & \texttt{float32} \\
    \midrule
    \multicolumn{2}{@{}l}{\emph{Ridge / Lasso}} \\
    Ridge $\alpha$              & $1.0$ \\
    Lasso $\alpha$ / max iter   & $0.01$ / $5000$ \\
    \midrule
    \multicolumn{2}{@{}l}{\emph{SVR}} \\
    Kernel / $C$                & RBF / $1.0$ \\
    \midrule
    \multicolumn{2}{@{}l}{\emph{Mass-mean}} \\
    Class split                 & target median \\
    Direction                   & $\bm{\mu}_{+}-\bm{\mu}_{-}$, unit-norm \\
    Metric                      & Spearman $\rho$ (uncalibrated) \\
    \midrule
    \multicolumn{2}{@{}l}{\emph{Logistic regression}} \\
    Target binning              & median split ($2$ classes) \\
    Inverse reg.\ $C$ / max iter & $1.0$ / $1000$ \\
    Class weighting             & uniform \\
    Metric                      & accuracy / $F_1$ / ROC-AUC \\
    \midrule
    \multicolumn{2}{@{}l}{\emph{MLP}} \\
    Hidden layer / dropout      & $1\times256$, ReLU / $0.1$ \\
    Optimiser / lr / wd         & Adam / $10^{-3}$ / $10^{-4}$ \\
    Batch size                  & $64$ \\
    Max.\ epochs / patience     & $100$ / $10$ \\
    Loss                        & MSE \\
    \bottomrule
  \end{tabular}
  \caption{Probe hyperparameters. Ridge, lasso, mass-mean and logistic are
  linear probes over standardised activations; SVR and the MLP are nonlinear.
  All share the same $80/20$ seed-$42$ split per (space, target).}
  \label{tab:probe-config}
\end{table}

\begin{table}[t]
  \centering
  \small
  \setlength{\tabcolsep}{4pt}
  \begin{tabular}{@{}lcccc@{}}
    \toprule
    \textbf{Norm} & \textbf{Ridge} & \textbf{SVR} & \textbf{MLP} & \textbf{Mass-mean} \\
     & $R^2$ & $R^2$ & $R^2$ & $\rho$ \\
    \midrule
    \multicolumn{5}{@{}l}{\emph{MiniLM-384}} \\
    concreteness & 0.752 & \textbf{0.756} & 0.711 & 0.714 \\
    imageability & 0.693 & \textbf{0.715} & 0.664 & 0.703 \\
    valence      & 0.587 & \textbf{0.678} & 0.656 & 0.671 \\
    aoa          & 0.607 & \textbf{0.647} & 0.578 & 0.661 \\
    semsize      & 0.532 & \textbf{0.595} & 0.568 & 0.616 \\
    gender       & 0.486 & \textbf{0.588} & 0.529 & 0.649 \\
    familiarity  & 0.519 & \textbf{0.552} & 0.502 & 0.598 \\
    arousal      & 0.434 & \textbf{0.511} & 0.462 & 0.540 \\
    dominance    & 0.396 & \textbf{0.502} & 0.445 & 0.534 \\
    \midrule
    \multicolumn{5}{@{}l}{\emph{EmbeddingGemma-768}} \\
    concreteness & 0.765 & \textbf{0.812} & 0.660 & 0.860 \\
    imageability & 0.692 & \textbf{0.756} & 0.594 & 0.809 \\
    valence      & 0.676 & \textbf{0.725} & 0.583 & 0.774 \\
    aoa          & 0.606 & \textbf{0.668} & 0.493 & 0.676 \\
    semsize      & 0.654 & \textbf{0.700} & 0.484 & 0.727 \\
    gender       & 0.533 & \textbf{0.613} & 0.347 & 0.651 \\
    familiarity  & 0.487 & \textbf{0.574} & 0.258 & 0.673 \\
    arousal      & 0.508 & \textbf{0.584} & 0.346 & 0.614 \\
    dominance    & 0.469 & \textbf{0.553} & 0.322 & 0.629 \\
    \bottomrule
  \end{tabular}
  \caption{Per-dimension recovery of the nine Glasgow norms ($4{,}682$ words),
  best-layer validation. Regression probes report $R^2$; the mass-mean direction
  reports Spearman $\rho$ (an uncalibrated projection, not comparable to the
  $R^2$ columns and therefore not bolded). SVR (bold) is the best regressor for
  every norm, and EmbeddingGemma beats MiniLM on every norm at its best probe.
  Norms are ordered by encodability.}
  \label{tab:glasgow-reg}
\end{table}

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{@{}lcccc@{}}
    \toprule
    & \multicolumn{2}{c}{\textbf{MiniLM-384}} & \multicolumn{2}{c}{\textbf{EmbGemma-768}} \\
    \cmidrule(lr){2-3}\cmidrule(lr){4-5}
    \textbf{Norm} & Acc & AUC & Acc & AUC \\
    \midrule
    concreteness & \textbf{0.871} & 0.953 & 0.838 & 0.922 \\
    imageability & 0.846 & 0.921 & 0.813 & 0.904 \\
    valence      & 0.790 & 0.870 & 0.801 & 0.877 \\
    aoa          & 0.795 & 0.877 & 0.777 & 0.848 \\
    semsize      & 0.793 & 0.873 & 0.794 & 0.874 \\
    gender       & 0.756 & 0.829 & 0.749 & 0.820 \\
    familiarity  & 0.809 & 0.881 & 0.758 & 0.835 \\
    arousal      & 0.741 & 0.815 & 0.730 & 0.804 \\
    dominance    & 0.712 & 0.793 & 0.710 & 0.775 \\
    \bottomrule
  \end{tabular}
  \caption{Logistic-regression probe (median split) on the nine Glasgow norms,
  best-layer validation accuracy and ROC-AUC. A linear classifier separates the
  above-/below-median halves of every norm well above chance ($0.5$), with
  concreteness the most separable ($0.87$ accuracy / $0.95$ AUC on MiniLM).}
  \label{tab:glasgow-logreg}
\end{table}

\paragraph{Concreteness at scale.}
Table~\ref{tab:conc-only} extends the concreteness probe to the full Brysbaert
set of $39{,}954$ words, using the final pooled MiniLM embedding (a single
vector, not the layer sweep). All five families recover concreteness strongly at
$8.5\times$ the vocabulary: the nonlinear probes lead (MLP $R^2=0.744$, SVR
$0.724$), the linear ridge direction reaches $R^2=0.697$, and a median-split
logistic classifier attains $0.842$ accuracy / $0.926$ AUC. Scores run slightly
below the $4{,}682$-word Glasgow figures, consistent with the larger, more
diverse vocabulary and the use of the pooled output rather than the best
internal layer.

\begin{table}[t]
  \centering
  \small
  \begin{tabular}{@{}lccccc@{}}
    \toprule
    & \textbf{Ridge} & \textbf{SVR} & \textbf{MLP} & \textbf{Mass-mean} & \textbf{Logistic} \\
    \midrule
    $R^2$        & 0.697 & 0.724 & \textbf{0.744} & 0.603 & --- \\
    $\rho$       & 0.835 & 0.846 & \textbf{0.856} & 0.802 & --- \\
    Acc / AUC    & ---   & ---   & ---   & ---   & 0.842 / 0.926 \\
    \bottomrule
  \end{tabular}
  \caption{Concreteness recovery on the full Brysbaert set ($39{,}954$ words),
  MiniLM final pooled embedding, seed-$42$ $80/20$ split. Mass-mean $R^2$ is the
  calibrated readout; the logistic probe uses a median split and reports
  accuracy/AUC.}
  \label{tab:conc-only}
\end{table}

\paragraph{In-platform variant.}
The same probe families are also exposed as an interactive ``find directions''
tool that trains on a collection's stored embedding vectors (read from the
vector store, no fresh model inference — a few seconds of CPU) against any
numeric metadata field, with a $5\times10^{4}$ training-subsample cap and full
scoring of all items. It reproduces the offline numbers: e.g.\ the in-platform
ridge probe on the $39{,}954$-word concreteness collection scores validation
$R^2=0.700$, matching Table~\ref{tab:conc-only}.
